---
title: "Lecture 15" 
subtitle: "GPs for GLMs + Spatial Data"
date: "10/26/2018"
fontsize: 11pt
output: 
  beamer_presentation:
    theme: metropolis
    highlight: pygments
    fig_caption: false
    keep_tex: true
    latex_engine: xelatex
    includes:
      in_header: ../settings.tex
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(patchwork)

set.seed(20180308)
```

```{r config, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width=7,
  fig.height=4.5,
  out.width="\\textwidth",
  fig.align="center",
  echo=TRUE,
  warning=FALSE
)

ggplot2::theme_set(ggplot2::theme_bw())

source("../util.R")
```


# GPs and GLMs

## Logistic Regression
 
A typical logistic regression problem uses the following model,

$$y_i \sim \text{Bern}(p_i)$$
$$\begin{aligned}
\text{logit}(p_i) 
  &= \symbf{X}\,\symbf{\beta} \\
  &= \beta_0 + \beta_1 \, x_{i1} + \cdots + \beta_k \, x_{ik}
\end{aligned}$$

. . .

there is no reason that the linear equation above can't contain thing like random effects or GPs

$$\begin{aligned}
y_i &\sim \text{Bern}(p_i) \\
\text{logit}(p_i) 
  &= \symbf{X}\,\symbf{\beta} + w(\symbf{x}) \\
\end{aligned}$$
where
$$ w(\symbf{x}) \sim \mathcal{N}(0,\Sigma)$$

## A toy example

```{r echo=FALSE}
n=200

eta = function(x) 2.5*sin(2.1*pi*(x-0.25)) 

d = data_frame(x=runif(n)) %>%
  mutate(
    eta = eta(x),
    p = inv_logit(eta(x)),
    y = rbinom(length(x), size=1, prob = inv_logit(eta(x)))
  )

(ggplot(d, aes(x=x, y=eta)) + geom_line() +
 ggplot(d, aes(x=x, y=p)) + geom_line()) / 
(ggplot(d, aes(x=x, y=y)) + geom_jitter(height = 0.05))
```


## Jags Model$^*$

\scriptoutput
```{r}
logistic_model = "model{
  for(i in 1:N) {
    y[i] ~ dbern(p[i])
    logit(p[i]) = beta0 + eta[i]
  }

  eta ~ dmnorm(rep(0,N), inverse(Sigma))

  for (i in 1:(length(y)-1)) {
    for (j in (i+1):length(y)) {
      Sigma[i,j] = sigma2 * exp(- l * d[i,j]))
      Sigma[j,i] = Sigma[i,j]
    }
  }

  for (i in 1:length(y)) {
    Sigma[i,i] = sigma2
  }  
  
  beta0 ~ dnorm(0, 1)
  sigma2 = 1/tau
  tau ~ dgamma(1, 2)
  l ~ dunif(3/0.5, 3/0.01)
}"
```

## Model Results - Diagnostics

```{r fit_log, echo=FALSE}
y = d$y
x = d$x

coords = cbind(x, 0)
n = nrow(coords)

n.batch = 200
batch.length = 100

n.samples = n.batch * batch.length
burnin = n.samples*0.5 + 1
thin = 10

if (!file.exists("logistic_model.rds")) {
  m = spBayes::spGLM(
    y~1, family="binomial", coords=coords,
    starting=list("beta"=0, "phi"=3/0.053,"sigma.sq"=1, "w"=0),
    tuning=list("beta"=0.5, "phi"=0.5, "sigma.sq"=0.5, "w"=0.5),
    priors=list("beta.Normal"=list(0,1), "phi.Unif"=c(3/0.5, 3/0.01),   "sigma.sq.IG"=c(2, 1)),
    amcmc=list("n.batch"=n.batch, "batch.length"=batch.length, "accept.rate"=0.43),
    cov.model="exponential", verbose=TRUE, n.report=10)
  
  saveRDS(m, file="logistic_model.rds")
} else {
  m = readRDS("logistic_model.rds")
}

mcmc = m$p.beta.theta.samples %>% 
  window(start = burnin, thin = thin) %>%
  tidybayes::gather_samples(`(Intercept)`, phi, sigma.sq)

ggplot(mcmc, aes(x=.iteration, y=estimate, color=term)) +
  geom_line() +
  facet_grid(term~., scales = "free_y") +
  guides(color=FALSE)
```


## Model Results - Fit

```{r make_pred, echo=FALSE}
pred = spBayes::spPredict(m, pred.coords = coords, pred.covars = matrix(rep(1,n)), start = burnin, thin=thin, verbose = FALSE)

pred = cbind(
  t(pred$p.w.predictive.samples),
  t(pred$p.y.predictive.samples)
) 
colnames(pred) = c(paste0("eta[",1:n,"]"), paste0("p[",1:n,"]"))

tidybayes::gather_samples(pred, eta[i],p[i]) %>%
  full_join(
    select(d, x) %>% mutate(i = 1:n()),
    by = "i"
  ) %>%
    ungroup() %>%
    ggplot(aes(x=x, y=estimate)) +
      tidybayes::stat_lineribbon(alpha=0.5) +
      geom_line(data=tidyr::gather(d, term, estimate, -x, -y), size=2, linetype=2) +
      facet_wrap(~term, scales = "free_y")
```

## Model vs glm

```{r}
g = glm(y~poly(x,2), data=d, family="binomial")
```

```{r echo=FALSE}
rbind(
  mutate(d, type = "truth"),
  mutate(d, 
    type = "glm prediction",
    eta = predict(g, type="link"),
    p = predict(g, type="response")
  )
) %>% 
  tidyr::gather(term, estimate, -x, -y, -type) %>%
  ggplot(aes(x=x, y=estimate, color=type, linetype=type)) +
    geom_line(size=2) +
    facet_wrap(~term, scales = "free_y") +
    scale_color_manual(values = c("red","black"))
```


## Count data - Polio cases

`Polio` from the glarma package.

$~$

> This data set gives the monthly number of cases of poliomyelitis in the U.S. for the years 1970–1983 as reported by the Center for Disease Control.


```{r echo=FALSE, fig.height=3.5}
data(Polio, package="glarma")

polio = data_frame(
  year = seq(1970, 1984-1/12, by=1/12),
  cases = Polio$Cases
)

ggplot(polio, aes(x=year, y=cases)) +
  geom_line() +
  geom_point()
```
 
## Polio Model

**Model**:
$$\begin{aligned}
y_i &\sim \text{Pois}(\lambda_i) \\
\text{log}(\lambda_i) 
  &= \beta_0 + w(\symbf{t}) \\
\end{aligned}$$
where
$$ w(\symbf{t}) \sim \mathcal{N}(0,\Sigma) $$
$$\{\symbf{\Sigma}\}_{ij} = \sigma^2 \exp(-|\phi \, d_{ij}|)$$

**Priors**:

$$
\begin{aligned}
\beta_0 &\sim \mathcal{N}(0,1) \\
\phi &\sim \text{Unif}\left(\frac{3}{6}, \frac{3}{1/12}\right)\\
1/\sigma^2 &\sim \text{Gamma}(2,1)
\end{aligned}
$$

## Model Results - Diagnostics

```{r echo=FALSE}
coords = cbind(polio$year, 0)
n = nrow(coords)

n.batch = 200
batch.length = 100

n.samples = n.batch * batch.length
burnin = n.samples*0.5 + 1
thin = 10

if (!file.exists("pois_model.rds")) {
  m = spBayes::spGLM(
    cases~1, data=polio, family="poisson", coords=coords,
    starting=list("beta"=0, "phi"=3/2,"sigma.sq"=1, "w"=0),
    tuning=list("beta"=0.5, "phi"=0.5, "sigma.sq"=0.5, "w"=0.5),
    priors=list("beta.Normal"=list(0,1), "phi.Unif"=c(3/6, 3/(1/12)),   "sigma.sq.IG"=c(2, 1)),
    amcmc=list("n.batch"=n.batch, "batch.length"=batch.length, "accept.rate"=0.43),
    cov.model="exponential", verbose=TRUE, n.report=10)
  
   m = spBayes::spGLM(
    cases~1, data=polio, family="poisson", coords=coords,
    starting=list("beta"=0, "phi"=sqrt(3)/2,"sigma.sq"=1, "w"=0),
    tuning=list("beta"=0.5, "phi"=0.5, "sigma.sq"=0.5, "w"=0.5),
    priors=list("beta.Normal"=list(0,1), "phi.Unif"=c(sqrt(3)/6, sqrt(3)/(1/12)),   "sigma.sq.IG"=c(2, 1)),
    amcmc=list("n.batch"=n.batch, "batch.length"=batch.length, "accept.rate"=0.43),
    cov.model="gaussian", verbose=TRUE, n.report=10)
  
  saveRDS(m, file="pois_model.rds")
} else {
  m = readRDS("pois_model.rds")
}

mcmc = m$p.beta.theta.samples %>% 
  window(start = burnin, thin = thin) %>%
  tidybayes::gather_samples(`(Intercept)`, phi, sigma.sq)

ggplot(mcmc, aes(x=.iteration, y=estimate, color=term)) +
  geom_line() +
  facet_grid(term~., scales = "free_y") +
  guides(color=FALSE)
```

## Model Results - Fit

```{r echo=FALSE}
pred = spBayes::spPredict(m, pred.coords = coords, pred.covars = matrix(rep(1,n)), start = burnin, thin=thin, verbose = FALSE)

pred = t(pred$p.y.predictive.samples)
colnames(pred) = paste0("lambda[",1:n,"]")

tidybayes::gather_samples(pred, lambda[i]) %>%
  full_join(
    mutate(polio, i = 1:n()),
    by = "i"
  ) %>%
    ungroup() %>%
    ggplot(aes(x=year, y=estimate)) +
      tidybayes::stat_lineribbon(alpha=0.5) +
      geom_point(data=polio, aes(y=cases)) +
      facet_wrap(~term, scales = "free_y")
```



# Spatial data in R

```{r include=FALSE}
library(sf)
library(stringr)
library(purrr)
```


## Analysis of geospatial data in R {.t}

R has a rich package ecosystem for read/writing, manipulating, and analyzing geospatial data.

\vspace{2mm}

Some core packages (CRAN - [Spatial task view](http://cran.r-project.org/web/views/Spatial.html)):

* `sp` - core classes for handling spatial data, additional utility functions.

* `rgdal` - R interface to `gdal` (Geospatial Data Abstraction Library) for reading and writing spatial data.

* `rgeos` - R interface to `geos` (Geometry Engine Open Source) library for querying and manipulating spatial data. Reading and writing WKT.

* `sf` - Combines the functionality of `sp`, `rgdal`, and `rgeos` into a single package based on tidy priciples.

* `lwgeom` - additional functionality for `sf` using PostGIS' liblwgeom.

* `raster` - classes and tools for handling raster data.



## Installing `sf` {.t}

This is the hardest part of using the `sf` package, difficulty comes from is dependence on several external libraries (`geos`, `gdal`, and `proj`).

* *Windows* - installing from source works when Rtools is installed (system requirements are downloaded from rwinlib)

* *MacOS* - install dependencies via homebrew: `gdal2`, `geos`, `proj`.

* *Linux* - Install development pacakages for GDAL (>= 2.0.0), GEOS (>= 3.3.0) and Proj.4 (>= 4.8.0) from your package manager of choice.


More specific details are included in the [repo readme on github](https://github.com/r-spatial/sf).



## Simple Features

```{r, echo=FALSE}
par(mar=c(1,1,2,1), mfrow=c(2,4))

## Single Geometries

pt = st_point(c(30, 10))
ls = st_linestring(matrix(c(30, 10, 10, 30, 40, 40), byrow=TRUE, ncol=2))
poly = st_polygon(list(matrix(c(30, 10, 40, 40, 20, 40, 10, 20, 30, 10), ncol=2, byrow=TRUE)))
poly_hole = st_polygon(
              list(
                matrix(c(35, 10, 45, 45, 15, 40, 10, 20, 35, 10), ncol=2, byrow=TRUE),
                matrix(c(20, 30, 35, 35, 30, 20, 20, 30), ncol=2, byrow=TRUE)
              )
            )

plot(pt, axes=FALSE, main="Point", pch=16); box()
plot(ls, axes=FALSE, main="Linestring");    box()
plot(poly, axes=FALSE, col="lightgrey", main="Polygon");  box()
plot(poly_hole, axes=FALSE, col="lightgrey", main="Polygon w/ Hole(s)");  box()


## Multi Geometries

pts = st_multipoint(matrix(c(10, 40, 40, 30, 20, 20, 30, 10), ncol=2, byrow=TRUE))
lss = st_multilinestring(list( 
        matrix(c(10, 10, 20, 20, 10, 40), ncol=2, byrow=TRUE),
        matrix(c(40, 40, 30, 30, 40, 20, 30, 10), ncol=2, byrow=TRUE)
      ))
        
polys = st_multipolygon(list(
          list(matrix(c(30, 20, 45, 40, 10, 40, 30, 20), ncol=2, byrow=TRUE)),
          list(matrix(c(15, 5, 40, 10, 10, 20, 5, 10, 15, 5), ncol=2, byrow=TRUE))
        ))
polys_hole = st_multipolygon(list(
                list(matrix(c(40, 40, 20, 45, 45, 30, 40, 40), ncol=2, byrow=TRUE)),
                list(
                  matrix(c(20, 35, 10, 30, 10, 10, 30, 5, 45, 20, 20, 35), ncol=2, byrow=TRUE),
                  matrix(c(30, 20, 20, 15, 20, 25, 30, 20), ncol=2, byrow=TRUE)
                )
              ))

plot(pts, axes=FALSE, main="Multipoint", pch=16); box()
plot(lss, axes=FALSE, main="Multilinestring");    box()
plot(polys, axes=FALSE, col="lightgrey", main="Multipolygon");  box()
plot(polys_hole, axes=FALSE, col="lightgrey", main="Multipolygon w/ Hole(s)");  box()
```


## Reading, writing, and converting simple features {.t}

- ~~`maptools`~~
    * ~~`readShapePoints` / `writeShapePoints` - Shapefile w/ points~~
    * ~~`readShapeLines` / `writeShapeLines` - Shapefile w/ lines~~
    * ~~`readShapePoly` / `writeShapePoly` - Shapefile w/ polygons~~
    * ~~`readShapeSpatial` / `writeShapeSpatial` - Shapefile~~
- ~~`rgdal`~~
    * ~~`readOGR` / `writeOGR` - Shapefile, GeoJSON, KML, ...~~
- `rgeos`
    * ~~`readWKT` / `writeWKT` - Well Known Text~~
- `sf`
    * `st_read` / `st_write` - Shapefile, GeoJSON, KML, ...
    * `st_as_sfc` / `st_as_wkt` - WKT
    * `st_as_sfc` / `st_as_binary` - WKB
    * `st_as_sfc` / `as(x, "Spatial")` - sp

See [sf vignette #2](https://cran.r-project.org/web/packages/sf/vignettes/sf2.html).

# Geospatial data in the real world

## Projections

```{r projs, echo=FALSE, message=FALSE}
lat_lines  = map(seq(9.999, 89.999, length.out = 9), ~ cbind(seq(-179.999, -9.999, length.out=100), .))
long_lines = map(seq(-179.999, -9.999, length.out = 17), ~ cbind(., seq(9.999, 89.999, length.out=100))) 

lat_long = c(lat_lines, long_lines) %>% 
  st_multilinestring() %>%
  st_sfc() %>%
  st_set_crs("+proj=longlat +datum=WGS84 +no_defs")

data(wrld_simpl, package = "maptools")
world = st_as_sf(wrld_simpl)

NAm = world %>% filter(FIPS %in% c("CA","GL","MX","US"))
NAm_google = st_transform(NAm, "+init=epsg:3857")


par(mar=c(3,2,2,1),mfrow=c(2,3))
plot(lat_long, col=adjustcolor("grey",alpha.f = 0.5), axes=TRUE, main="Lat/Long (epsg:4326)")
plot(st_geometry(NAm), col="black", add=TRUE)

plot(st_transform(lat_long, "+init=epsg:3857"), col=adjustcolor("grey",alpha.f = 0.5), axes=TRUE, main="Google / Web Mercator (epsg:3857)", ylim=c(0, 2e7))
plot(st_transform(NAm, "+init=epsg:3857") %>% st_geometry(), col="black", add=TRUE)

lcc = "+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
plot(st_transform(lat_long, lcc), col=adjustcolor("grey",alpha.f = 0.5), axes=TRUE, main="Lambert Conformal Conic:")
plot(st_transform(NAm, lcc) %>% st_geometry(), col="black", add=TRUE)

aea = "+proj=aea +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
plot(st_transform(lat_long, aea), col=adjustcolor("grey",alpha.f = 0.5), axes=TRUE, main="Alberts Equal Area")
plot(st_transform(NAm, aea) %>% st_geometry(), col="black", add=TRUE)

robinson = "+proj=robin +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
plot(st_transform(lat_long, robinson), col=adjustcolor("grey",alpha.f = 0.5), axes=TRUE, main="Robinson")
plot(st_transform(NAm, robinson) %>% st_geometry(), col="black", add=TRUE)

mollweide = "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
plot(st_transform(lat_long, mollweide), col=adjustcolor("grey",alpha.f = 0.5), axes=TRUE, main="Mollweide")
plot(st_transform(NAm, mollweide) %>% st_geometry(), col="black", add=TRUE)
```

## Distance on a Sphere

```{r echo=FALSE}
states = st_read("../data/us/states-unfiltered.shp", quiet = TRUE, stringsAsFactors = FALSE) %>%
  filter(!STATE %in% c("Alaska", "Hawaii", "Puerto Rico", "U.S. Virgin Islands"))

plot(st_geometry(states))

durham = c(-78.8986, 35.9940)
la = c(-118.243, 34.0522)
locs = rbind(durham, la)

gc = geosphere::gcIntermediate(durham, la, n=50, addStartEnd=TRUE)

points(locs, col='blue', pch=16, cex=2)
lines(locs, col='red', lty=2, lwd=2)
lines(gc, col='red', lty=3, lwd=2)
```

## Dateline

Want to fly from the Western most point in the US to the Eastern most point?

```{r echo=FALSE, message=FALSE, fig.height=5, fig.width=8}
par(mar=c(3,3,1,1))

ak = st_read("../data/ak/states.shp", quiet = TRUE, stringsAsFactors = FALSE)
ak_geom = st_geometry(ak)

west_hem = st_polygon(list(matrix(c(-180,90, -180,-90, 0,-90, 0,90, -180,90), ncol=2,byrow=TRUE))) %>% 
  st_sfc() %>%
  st_set_crs("+proj=longlat +datum=WGS84")

east_hem = st_polygon(list(matrix(c(180,90, 180,-90, 0,-90, 0,90, 180,90), ncol=2,byrow=TRUE))) %>% 
  st_sfc() %>%
  st_set_crs("+proj=longlat +datum=WGS84")

ak_west = st_intersection(ak_geom, west_hem)
ak_east = st_intersection(ak_geom, east_hem)

ak_east_shift = (ak_east - c(360,0)) %>% st_set_crs("+proj=longlat +datum=WGS84")

ak_shift = st_union(ak_east_shift, ak_west)

plot(ak_shift, axes=TRUE, col="black", border=NA, xlim=c(-190, -130))
points(c(-360+179.776,-179.146), c(51.952,51.273),col='red')
abline(v=-180,col='blue',lty=2)
```

##

```{r echo=FALSE}
plot(ak_shift, axes=TRUE, col="black", border=NA, xlim=c(-190, -170), ylim=c(50, 55))
points(c(-360+179.776,-179.146), c(51.952,51.273),col='red')
abline(v=-180,col='blue',lty=2)
```


##

```{r echo=FALSE, fig.align="center", fig.width=8, fig.height=4}
library(geosphere)
par(mar=c(0,0,0,0))
inter = gcIntermediate(c(179.776,51.952), c(-179.146,51.273), n=50, addStartEnd=TRUE)

plot(st_geometry(world), col="black", ylim=c(-90,90), axes=TRUE)
lines(inter,col='red',lwd=2,lty=3)
```





# Using `sf`

## Example data {.t}

\scriptoutput

```{r}
nc  = st_read("../data/gis/nc_counties/", quiet=TRUE, stringsAsFactors=FALSE)
air = st_read("../data/gis/airports/", quiet=TRUE, stringsAsFactors=FALSE)
hwy = st_read("../data/gis/us_interstates/", quiet=TRUE, stringsAsFactors=FALSE)

tbl_df(nc)
```

## {.t}

\scriptoutput

```{r}
tbl_df(air)
```

## {.t}

\scriptoutput

```{r}
tbl_df(hwy)
```

## `sf` classes

\scriptoutput

```{r}
str(nc)

class(nc)

class(nc$geometry)

class(nc$geometry[[1]])
```


## Projections {.t}

```{r}
st_crs(nc)
st_crs(air)
st_crs(hwy)
```

```{r echo=FALSE, fig.align="center", fig.height=3.5, fig.width=8}
par(mar=c(3,3,3,0.1), mfrow=c(1,3))
plot(st_geometry(nc),  axes=TRUE, main="nc")
plot(st_geometry(air), axes=TRUE, pch=16, col="blue", main="air")
plot(st_geometry(hwy), axes=TRUE, col="red", main="hwy")
```

## UTM Zones

\begin{center}
\includegraphics[width=0.9\textwidth]{figs/UTM_Zones.png}
\end{center}

## Lat/Long

```{r}
nc_ll = nc
air_ll = air
hwy_ll = st_transform(hwy, st_crs(nc))
```

```{r echo=FALSE, fig.align="center", fig.height=3.5, fig.width=8}
par(mar=c(3,3,3,0.1), mfrow=c(1,3))
plot(st_geometry(nc_ll),  axes=TRUE, main="nc")
plot(st_geometry(air_ll), axes=TRUE, pch=16, col="blue", main="air")
plot(st_geometry(hwy_ll), axes=TRUE, col="red", main="hwy")
```


## UTM

```{r}
nc_utm = st_transform(nc, st_crs(hwy))
air_utm = st_transform(air, st_crs(hwy))
hwy_utm = hwy
```


```{r echo=FALSE, fig.align="center", fig.height=3.5, fig.width=8}
par(mar=c(3,3,3,0.1), mfrow=c(1,3))
plot(st_geometry(nc_utm),  axes=TRUE, main="nc")
plot(st_geometry(air_utm), axes=TRUE, pch=16, col="blue", main="air")
plot(st_geometry(hwy_utm), axes=TRUE, col="red", main="hwy")
```

## Comparison

```{r echo=FALSE, fig.align="center", fig.height=3}
par(mar=c(3,3,3,0.1), mfrow=c(1,2))

plot(st_geometry(nc_ll),  axes=TRUE, main="Lat/Long")
plot(st_geometry(hwy_ll), col="red", add=TRUE)
plot(st_geometry(air_ll), pch=16, col="blue", add=TRUE)

plot(st_geometry(nc_utm),  axes=TRUE, main="UTM")
plot(st_geometry(hwy_utm), col="red", add=TRUE)
plot(st_geometry(air_utm), pch=16, col="blue", add=TRUE)
```

# Geometry Predicates

## DE-9IM

\begin{center}
\includegraphics[width=0.95\textwidth]{figs/de_9im.png}
\end{center}

## Spatial predicates 

\begin{center}
\includegraphics[width=0.75\textwidth]{figs/predicates.png}
\end{center}

\footnotesize


\hspace*{-4pt}\makebox[\linewidth][c]{
\begin{tabular}{lll}

st\_within(a,b) & & st\_touches(a,b) \\

$\begin{bmatrix} T & * & F \\ * & * & F \\ * & * & * \end{bmatrix}$ & &
$\begin{bmatrix} F & T & * \\ * & * & * \\ * & * & * \end{bmatrix} \cup 
\begin{bmatrix} F & * & * \\ T & * & * \\ * & * & * \end{bmatrix} \cup 
\begin{bmatrix} F & * & * \\ * & T & * \\ * & * & * \end{bmatrix}$ \\

\\

st\_crosses(a,b) & & st\_overlaps(a,b) ($\text{dim}(a) = \text{dim}(b)$) \\

$\overset{\text{If dim}(a) < \text{dim}(b)}{\begin{bmatrix} T & * & T \\ * & * & * \\ * & * & * \end{bmatrix}} ~~~
\overset{\text{If dim}(a) > \text{dim}(b)}{\begin{bmatrix} T & * & * \\ * & * & * \\ T & * & * \end{bmatrix}} ~~~
\overset{\text{If dim}(any) = 1}{\begin{bmatrix} 0 & * & * \\ * & * & * \\ * & * & * \end{bmatrix}}$ & &
$\overset{\text{If dim} \in \{0,2\}}{\begin{bmatrix} T & * & T \\ * & * & * \\ T & * & * \end{bmatrix}} ~~~
\overset{\text{If dim} = 1}{\begin{bmatrix} 1 & * & T \\ * & * & * \\ T & * & * \end{bmatrix}}$ \\
\end{tabular}
}


## Sparse vs Full Results

\scriptsize

```{r}
st_intersects(nc[20:30,], air) %>% str()
```

##

```{r}
st_intersects(nc, air, sparse=FALSE) %>% str()
```

## Examples {.t}

* Which counties are adjacent to Durham County?

* Which counties have more than 4 neighbors?

* Which counties have an airport?