---
title: "Lecture 10" 
subtitle: "Seasonal Arima"
date: "10/05/2018"
fontsize: 11pt
output: 
  beamer_presentation:
    theme: metropolis
    highlight: pygments
    fig_caption: false
    keep_tex: true
    latex_engine: xelatex
    includes:
      in_header: ../settings.tex
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)

set.seed(20180222)
```

```{r config, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width=7,
  fig.height=4.5,
  out.width="\\textwidth",
  fig.align="center",
  echo=TRUE,
  warning=FALSE
)

ggplot2::theme_set(ggplot2::theme_bw())

source("../util.R")
```

---
class: middle
count: false

# Seasonal Models

---

## Australian Wine Sales Example (Lecture 6)

Australian total wine sales by wine makers in bottles <= 1 litre. Jan 1980 – Aug 1994.

```{r echo=FALSE}
data(wineind, package = "forecast")
forecast::ggtsdisplay(wineind, points = FALSE)
```

---

## Differencing

```{r echo=FALSE}
forecast::ggtsdisplay(diff(wineind), points = FALSE)
```

---

## Seasonal Arima {.t}

We can extend the existing Arima model to handle these higher order lags (without having to include all of the intervening lags).

\vspace{3mm}

Seasonal $\text{ARIMA}\,(p,d,q) \times (P,D,Q)_s$:
$$ \Phi_P(L^s) \, \phi_p(L) \, \Delta_s^D \, \Delta^d \, y_t = \delta + \Theta_Q(L^s) \, \theta_q(L) \, w_t$$
\pause where
$$
\begin{aligned}
\phi_p(L) &= 1-\phi_1 L - \phi_2 L^2 - \ldots - \phi_p L^p\\
\theta_q(L) &= 1+\theta_1 L + \theta_2 L^2 + \ldots + \theta_p L^q \\
\Delta^d &= (1-L)^d\\
\\
\Phi_P(L^s) &= 1-\Phi_1 L^s - \Phi_2 L^{2s} - \ldots - \Phi_P L^{Ps} \\
\Theta_Q(L^s) &= 1+\Theta_1 L + \Theta_2 L^{2s} + \ldots + \theta_p L^{Qs} \\
\Delta_s^D &= (1-L^s)^D\\
\end{aligned}
$$

---

## Seasonal Arima for \texttt{wineind} - AR

Lets consider an $\text{ARIMA}(0,0,0) \times (1,0,0)_{12}$:
$$
\begin{aligned}
(1-\Phi_1 L^{12}) \, y_t = \delta + w_t \\
y_t = \Phi_1 y_{t-12} + \delta + w_t
\end{aligned}
$$
\vspace{2mm}

```{r}
(m1.1 = forecast::Arima(wineind, seasonal=list(order=c(1,0,0), period=12)))
```

---

## Fitted model

```{r echo=FALSE, fig.height=4}
cbPalette = c("#000000", "#ff3333", "#92c5de")

model = "Model 1.1 - Arima (0,0,0) x (1,0,0)[12]"
rmse = (wineind-m1.1$fitted)^2 %>% mean() %>% sqrt() %>% round(2) %>% paste0("      [RMSE: ", . ,"]")

m1.1 %>%
  {data_frame(
    wineind = wineind %>% strip_attrs(),
    model   = .$fitted %>% strip_attrs(),
    time    = time(.$fitted) %>% strip_attrs()
  )} %>% 
  tidyr::gather(var, sales, -time) %>%
  mutate(var = forcats::as_factor(var)) %>%
  ggplot(aes(x=time, y=sales, color=var)) + 
    geom_line(alpha=0.75, size=0.8) +
    scale_colour_manual(values=cbPalette, name="") +
    labs(title=paste(model, rmse))
```

---

## Seasonal Arima for \texttt{wineind} - Diff {.t}

Lets consider an $\text{ARIMA}(0,0,0) \times (0,1,0)_{12}$:
$$
\begin{aligned}
(1 - L^{12}) \, y_t = \delta + w_t \\
y_t = y_{t-12} + \delta + w_t
\end{aligned}
$$
\vspace{2mm}

```{r}
(m1.2 = forecast::Arima(wineind, seasonal=list(order=c(0,1,0), period=12)))
```


---

## Fitted model

```{r echo=FALSE, fig.height=4}
cbPalette = c("#000000", "#ff3333", "#92c5de")

model = "Model 1.2 - Arima (0,0,0) x (0,1,0)[12]"
rmse = (wineind-m1.2$fitted)^2 %>% mean() %>% sqrt() %>% round(2) %>% paste0("      [RMSE: ", . ,"]")

m1.2 %>%
  {data_frame(
    wineind = wineind %>% strip_attrs(),
    model   = .$fitted %>% strip_attrs(),
    time    = time(.$fitted) %>% strip_attrs()
  )} %>% 
  tidyr::gather(var, sales, -time) %>%
  mutate(var = forcats::as_factor(var)) %>%
  ggplot(aes(x=time, y=sales, color=var)) + 
    geom_line(alpha=0.75, size=0.8) +
    scale_colour_manual(values=cbPalette, name="") +
    labs(title=paste(model, rmse))
```

---

## Residuals - Model 1.1

```{r echo=FALSE}
forecast::ggtsdisplay(m1.1$residuals)
```


---

## Residuals - Model 1.2

```{r echo=FALSE}
forecast::ggtsdisplay(m1.2$residuals)
```


---

## Model 2

$\text{ARIMA}(0,0,0) \times (0,1,1)_{12}$:
$$
\begin{aligned}
(1-L^{12}) y_t = \delta + (1+\Theta_1 L^{12})  w_t \\
y_t-y_{t-12} = \delta + w_t + \Theta_1 w_{t-12} \\
y_t = \delta + y_{t-12} + w_t + \Theta_1 w_{t-12}
\end{aligned}
$$


```{r}
(m2 = forecast::Arima(wineind, order=c(0,0,0), 
                      seasonal=list(order=c(0,1,1), period=12)))
```

---

## Fitted model

```{r echo=FALSE, fig.height=4}
model = "Model 2 - forecast::Arima (0,0,0) x (0,1,1)[12]"
rmse = (wineind-m2$fitted)^2 %>% mean() %>% sqrt() %>% round(2) %>% paste0("      [RMSE: ", . ,"]")

m2 %>%
  {data_frame(
    wineind = wineind %>% strip_attrs(),
    model   = .$fitted %>% strip_attrs(),
    time    = time(.$fitted) %>% strip_attrs()
  )} %>% 
  tidyr::gather(var, sales, -time) %>%
  mutate(var = forcats::as_factor(var)) %>%
  ggplot(aes(x=time, y=sales, color=var)) + 
    geom_line(alpha=0.75, size=0.8) +
    scale_colour_manual(values=cbPalette, name="") +
    labs(title=paste(model, rmse))
```

---

## Residuals

```{r echo=FALSE}
forecast::ggtsdisplay(m2$residuals)
```


---

## Model 3

$\text{ARIMA}(3,0,0) \times (0,1,1)_{12}$
$$
\begin{aligned}
(1-\phi_1 L - \phi_2 L^2 - \phi_3 L^3) \, (1-L^{12}) y_t = \delta + (1 + \Theta_1 L)w_t \\
(1-\phi_1 L - \phi_2 L^2 - \phi_3 L^3) \, (y_t-y_{t-12}) = \delta + w_t + w_{t-12} \\
y_t = \delta + \sum_{i=1}^3 \phi_i y_{t-1}  + y_{t-12}  - \sum_{i=1}^3 \phi_i y_{t-12-i} + w_t + w_{t-12}
\end{aligned}
$$


```{r}
(m3 = forecast::Arima(wineind, order=c(3,0,0), 
                      seasonal=list(order=c(0,1,1), period=12)))
```

---

## Fitted model

```{r echo=FALSE, fig.height=4}
model = "Model 3 - forecast::Arima (3,0,0) x (0,1,1)[12]"
rmse = (wineind-m3$fitted)^2 %>% mean() %>% sqrt() %>% round(2) %>% paste0("      [RMSE: ", . ,"]")

m3 %>%
  {data_frame(
    wineind = wineind %>% strip_attrs(),
    model   = .$fitted %>% strip_attrs(),
    time    = time(.$fitted) %>% strip_attrs()
  )} %>% 
  tidyr::gather(var, sales, -time) %>%
  mutate(var = forcats::as_factor(var)) %>%
  ggplot(aes(x=time, y=sales, color=var)) + 
    geom_line(alpha=0.75, size=0.8) +
    scale_colour_manual(values=cbPalette, name="") +
    labs(title=paste(model, rmse))
```

---

## Model - Residuals

```{r echo=FALSE}
forecast::ggtsdisplay(m3$residuals)
```

---
class: middle
count: false

# Federal Reserve Board Production Index

---

## \texttt{prodn} from the \texttt{astsa} package

Monthly Federal Reserve Board Production Index (1948-1978)

```{r}
data(prodn, package="astsa"); forecast::ggtsdisplay(prodn, points = FALSE)
```

---

## Differencing

Based on the ACF it seems like standard differencing may be required

```{r echo=FALSE}
forecast::ggtsdisplay(diff(prodn))
```

---

## Differencing + Seasonal Differencing

Additional seasonal differencing also seems warranted

```{r}
(fr_m1 = forecast::Arima(prodn, order = c(0,1,0), 
            seasonal = list(order=c(0,0,0), period=12)))
```

```{r}
(fr_m2 = forecast::Arima(prodn, order = c(0,1,0), 
            seasonal = list(order=c(0,1,0), period=12)))
```


---

## Residuals

```{r echo=FALSE}
forecast::ggtsdisplay(fr_m2$residuals, points=FALSE, lag.max=36)
```

---

## Adding Seasonal MA

\scriptoutput

```{r}
(fr_m3.1 = forecast::Arima(prodn, order = c(0,1,0), 
            seasonal = list(order=c(0,1,1), period=12)))
```

```{r}
(fr_m3.2 = forecast::Arima(prodn, order = c(0,1,0), 
            seasonal = list(order=c(0,1,2), period=12)))
```

---

## Adding Seasonal MA (cont.) {.t}

\scriptoutput 

```{r}
(fr_m3.3 = forecast::Arima(prodn, order = c(0,1,0), 
            seasonal = list(order=c(0,1,3), period=12)))
```


---

## Residuals - Model 3.3

```{r echo=FALSE}
forecast::ggtsdisplay(fr_m3.3$residuals, points=FALSE, lag.max = 36)
```


---

## Adding AR

\scriptoutput

```{r}
(fr_m4.1 = forecast::Arima(prodn, order = c(1,1,0), 
            seasonal = list(order=c(0,1,3), period=12)))
```

```{r}
(fr_m4.2 = forecast::Arima(prodn, order = c(2,1,0), 
            seasonal = list(order=c(0,1,3), period=12)))
```

---

## Residuals - Model 4.1

```{r echo=FALSE}
forecast::ggtsdisplay(fr_m4.1$residuals, points=FALSE, lag.max = 36)
```

---

## Residuals - Model 4.2

```{r echo=FALSE}
forecast::ggtsdisplay(fr_m4.2$residuals, points=FALSE, lag.max = 36)
```

---

## Model Fit

```{r echo=FALSE, fig.height=4}
model = "Model 4.1 - forecast::Arima (1,1,0) x (0,1,3)[12]"
rmse = (prodn-fr_m4.1$fitted)^2 %>% mean() %>% sqrt() %>% round(3) %>% paste0("      [RMSE: ", . ,"]")

fr_m4.1 %>%
  {data_frame(
    prodn = prodn %>% strip_attrs(),
    model = .$fitted %>% strip_attrs(),
    time  = time(.$fitted) %>% strip_attrs()
  )} %>% 
  tidyr::gather(var, sales, -time) %>%
  mutate(var = forcats::as_factor(var)) %>%
  ggplot(aes(x=time, y=sales, color=var)) + 
    geom_line(alpha=0.75, size=0.8) +
    scale_colour_manual(values=cbPalette, name="") +
    labs(title=paste(model, rmse))
```

---

## Model Forecast

```{r}
forecast::forecast(fr_m4.1) %>% plot()
```

---

## Model Forecast (cont.)

```{r}
forecast::forecast(fr_m4.1) %>% plot(xlim=c(1975,1982))
```

---

## Model Forecast (cont.)

```{r}
forecast::forecast(fr_m4.1, 120) %>% plot()
```

---

## Model Forecast (cont.)

```{r}
forecast::forecast(fr_m4.1, 120) %>% plot(xlim=c(1975,1990))
```

---

## Exercise - Cortecosteroid Drug Sales

Monthly cortecosteroid drug sales in Australia from 1992 to 2008.

```{r fig.height=4}
data(h02, package="fpp")
forecast::ggtsdisplay(h02,points=FALSE)
```

---

# Forecasting

---

## Forecasting ARMA {.t}

Forecasts for stationary models necessarily revert to mean 

  * Remember, $E(y_t) \ne \delta$ but rather $\delta / (1 - \sum_{i=1}^p \phi_i)$.
  
  * Differenced models revert to trend (usually a line)
    
  * Why? AR gradually damp out, MA terms disappear

\vspace{5mm}
  
Like any other model, accuracy decreases as we extrapolate and the prediction interval expands rapidly

---

## One step ahead forecasting {.t}

Take a fitted ARMA(1,1) process where we know $\delta$, $\phi$, and $\theta$ then

---

## ARIMA(3,1,1) example

