---
title: "Lecture 7" 
subtitle: "AR Models"
date: "9/26/2018"
fontsize: 11pt
output: 
  beamer_presentation:
    theme: metropolis
    highlight: pygments
    fig_caption: false
    latex_engine: xelatex
    includes:
      in_header: ../settings.tex
---
  
```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(forecast)
```

```{r config, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width=7,
  fig.height=4.5,
  out.width="\\textwidth",
  fig.align="center",
  echo=TRUE,
  warning=FALSE
)

ggplot2::theme_set(ggplot2::theme_bw())

source("../util.R")
```

# Lagged Predictors and CCFs

## Southern Oscillation Index & Recruitment

The Southern Oscillation Index (SOI) is an indicator of the development and intensity of El Niño (negative SOI) or La Niña (positive SOI) events in the Pacific Ocean. These data also included the estimate of "recruitment", which indicate fish population sizes in the southern hemisphere.

\scriptoutput

```{r echo=FALSE}
library(astsa)

fish = data_frame(
  date = time(astsa::soi) %>% strip_attrs(),
  soi = astsa::soi %>% strip_attrs(),
  recruitment = astsa::rec %>% strip_attrs()
)
fish
```

## Time series

```{r echo=FALSE}
fish %>%
  tidyr::gather(Variables, value, -date) %>%
  ggplot(aes(x=date, y=value, color=Variables)) +
    geom_line() +
    facet_wrap(~Variables, nrow=2, scale="free_y") +
    ylab("")
```

## Relationship?

```{r echo=FALSE, message=FALSE}
fish %>%
  ggplot(aes(x=soi, y=recruitment)) +
  geom_point() + 
  geom_smooth(color="red", alpha=0.5, se=FALSE)
```

## `soi`s ACF & PACF

```{r}
forecast::ggtsdisplay(fish$soi, lag.max = 36)
```

## `recruitment`

```{r}
forecast::ggtsdisplay(fish$recruitment, lag.max = 36)
```




## Cross correlation function

```{r}
with(fish, forecast::ggCcf(soi, recruitment))
```

## Cross correlation function - Scatter plots

```{r echo=FALSE, warning=FALSE, fig.align="center"}
fish_lags = fish %>%
  mutate(
    "lag 0" = lag(soi,0),
    "lag 1" = lag(soi,1),
    "lag 2" = lag(soi,2),
    "lag 3" = lag(soi,3),
    "lag 4" = lag(soi,4),
    "lag 5" = lag(soi,5),
    "lag 6" = lag(soi,6),
    "lag 7" = lag(soi,7),
    "lag 8" = lag(soi,8),
    "lag 9" = lag(soi,9),
    "lag 10" = lag(soi,10),
    "lag 11" = lag(soi,11)
  ) 

corrs = fish_lags %>%
  select(-soi, -date) %>%
  tidyr::gather(lag, soi, -recruitment) %>%
  group_by(lag) %>%
  summarize(corr = cor(soi, recruitment, use="complete.obs"))

fish_lags %>%
  select(-soi, -date) %>%
  tidyr::gather(lag, soi, -recruitment) %>%
  ggplot(aes(x=soi, y=recruitment)) +
    geom_point(alpha=0.3) +
    facet_wrap(~forcats::as_factor(lag)) +
    #geom_line(stat="smooth", method="loess", color='red',  size=1.2, alpha=0.75) +
    geom_line(stat="smooth", method="lm",    color="blue", size=1.2, alpha=0.75) +
    geom_text(data=corrs, aes(x=-0.85, y=5, label=round(corr,3)), size=3)
```


## Model

\scriptoutput

```{r}
model1 = lm(recruitment~lag(soi,6), data=fish)
model2 = lm(recruitment~lag(soi,6)+lag(soi,7), data=fish)
model3 = lm(recruitment~lag(soi,5)+lag(soi,6)+lag(soi,7)+lag(soi,8), data=fish)

summary(model3)
```

## Prediction

```{r echo=FALSE, warning=FALSE, fig.align="center"}
fish %>%
  modelr::add_predictions(., model1, var = paste0("Model 1 - soi lag 6 (RMSE: ",        round(modelr::rmse(model1,.),1),")")) %>%
  modelr::add_predictions(., model2, var = paste0("Model 2 - soi lags 6,7 (RMSE: ",     round(modelr::rmse(model2,.),1),")")) %>%
  modelr::add_predictions(., model3, var = paste0("Model 3 - soi lags 5,6,7,8 (RMSE: ", round(modelr::rmse(model3,.),1),")")) %>%
  tidyr::gather(model, pred, -(date:recruitment)) %>%
  ggplot(aes(x=date,y=recruitment)) +
    geom_line() +
    geom_line(aes(y=pred), col="red") +
    facet_wrap(~model,nrow=3)
```

## Residual ACF - Model 3

```{r echo=FALSE}
forecast::ggtsdisplay(residuals(model3))
```

## Autoregessive model 1

\scriptoutput

```{r}
model4 = lm(recruitment~lag(recruitment,1) + lag(recruitment,2) + 
                        lag(soi,5)+lag(soi,6)+lag(soi,7)+lag(soi,8), 
            data=fish)
summary(model4)
```

## Autoregessive model 2

\scriptoutput

```{r}
model5 = lm(recruitment~lag(recruitment,1) + lag(recruitment,2) + 
                        lag(soi,5) + lag(soi,6),
            data=fish)
summary(model5)
```

## Prediction

```{r echo=FALSE, warning=FALSE, fig.align="center"}
fish %>%
  modelr::add_predictions(., model3, var = paste0("Model 3 - soi lags 5,6,7,8 (RMSE: ",        round(modelr::rmse(model3,.),2),")")) %>%
  modelr::add_predictions(., model4, var = paste0("Model 4 - AR(2); soi lags 5,6,7,8 (RMSE: ", round(modelr::rmse(model4,.),2),")")) %>%
  modelr::add_predictions(., model5, var = paste0("Model 5 - AR(2); soi lags 5,6 (RMSE: ",     round(modelr::rmse(model5,.),2),")")) %>%
  tidyr::gather(model, pred, -(date:recruitment)) %>%
  ggplot(aes(x=date,y=recruitment)) +
    geom_line() +
    geom_line(aes(y=pred), col="red") +
    facet_wrap(~model,nrow=3)
```

## Residual ACF - Model 5

```{r}
forecast::ggtsdisplay(residuals(model5))
```

# Non-stationarity

## Non-stationary models {.t}

\vspace{5mm}

> All happy families are alike; each unhappy family is unhappy in its own way. 
> 
> - Tolstoy, Anna Karenina

This applies to time series models as well, just replace happy family with stationary model.

. . .


\vspace{3mm}

A simple example of a non-stationary time series is a trend stationary model 

$$ y_t = \mu(t) + w_t $$

where $\mu(t)$ denotes a time dependent trend and $w_t$ is a white noise (stationary) process. 


## Linear trend model

Lets imagine a simple model where $y_t = \delta + \beta t + x_t$ where $\delta$ and $\beta$ are constants and $x_t$ is a stationary process.

```{r echo=FALSE}
delta = 1
beta = 0.1
lt = data_frame(
  t = 1:100,
  w = rnorm(100)
) %>%
  mutate(y = delta + beta * t + w)

ggplot(lt, aes(x=t,y=y)) + 
  geom_point() +
  geom_line() + 
  labs(title="Linear trend")
```



## Differencing {.t}

An simple approach to remove trend is to difference your response variable, specifically examine $y_t - y_{t-1}$ instead of $y_t$.


## Detrending vs Difference

```{r echo=FALSE}
gridExtra::grid.arrange(
  lt %>%
    modelr::add_residuals(.,lm(y~t,data=.)) %>%
    ggplot(aes(x=t,y=resid)) + 
      geom_point() +
      geom_line() + 
      labs(title="Detrended"),
  data_frame(t=1:99, y_diff=diff(lt$y)) %>%
    ggplot(aes(x=t, y=y_diff)) +
      geom_point() +
      geom_line() + 
      labs(title="Differenced")
)
```

## Quadratic trend model

Lets imagine another simple model where $y_t = \delta + \beta t + \gamma t^2 + x_t$ where $\delta$, $\beta$, and $\gamma$ are constants and $x_t$ is a stationary process.

```{r echo=FALSE}
delta = -1
phi = -0.10
gamma = 0.0012
qt = data_frame(
  t = 1:100,
  w = rnorm(100)
) %>%
  mutate(y = ((t-50)/20)^2 + w)

ggplot(qt, aes(x=t,y=y)) + 
  geom_point() +
  geom_line() + 
  labs(title="Quadratic trend")
```

## Detrending

```{r echo=FALSE}
gridExtra::grid.arrange(
  qt %>%
    modelr::add_residuals(.,lm(y~t,data=.)) %>%
    ggplot(aes(x=t,y=resid)) + 
      geom_point() +
      geom_line() + 
      labs(title="Detrended - Linear"),
  qt %>%
    modelr::add_residuals(.,lm(y~t+I(t^2),data=.)) %>%
    ggplot(aes(x=t,y=resid)) + 
      geom_point() +
      geom_line() + 
      labs(title="Detrended - Quadratic")
)
```

## 2nd order differencing {.t}

Let $d_t = y_t - y_{t-1}$ be a first order difference then $d_t - d_{t-1}$ is a 2nd order difference.



## Differencing

```{r echo=FALSE}
gridExtra::grid.arrange(
  data_frame(t=1:99, y_diff=diff(qt$y)) %>%
    ggplot(aes(x=t, y=y_diff)) +
      geom_point() +
      geom_line() + 
      labs(title="1st Difference"),
  data_frame(t=1:98, y_diff=diff(qt$y, differences=2)) %>%
    ggplot(aes(x=t, y=y_diff)) +
      geom_point() +
      geom_line() + 
      labs(title="2nd Difference")
)
```

## Differencing - ACF

```{r echo=FALSE, fig.align="center"}
with(qt,
  gridExtra::grid.arrange(
    forecast::ggAcf(y),
    forecast::ggPacf(y),
    forecast::ggAcf(diff(y)),
    forecast::ggPacf(diff(y)),
    forecast::ggAcf(diff(y, differences = 2)),
    forecast::ggPacf(diff(y, differences = 2))
  )
)
```

# AR Models

## AR(1) {.t}

Last time we mentioned a random walk with trend process where $y_t = \delta + y_{t-1} + w_t$. 

The AR(1) process is a generalization of this where we include a coefficient in front of the $y_{t-1}$ term. 

$$AR(1): \quad y_t = \delta + \phi \, y_{t-1} + w_t $$
$$ w_t \sim N(0, \sigma^2_w) $$

## AR(1) - Positive $\phi$

```{r echo=FALSE, fig.height=5}
delta = 0.1
phi1 = 0.9
phi2 = 1.01

ar1 = data_frame(
  t = 1:500,
  y1 = 0,
  y2 = 0,
  y3 = 0
)

for(t in 2:nrow(ar1))
{
  ar1$y1[t] = delta + phi1 * ar1$y1[t-1] + rnorm(1)
  ar1$y2[t] = delta + ar1$y2[t-1] + rnorm(1)
  ar1$y3[t] = delta + phi2 * ar1$y3[t-1] + rnorm(1)
}

ar1 %>%
  rename(
    "AR(1) w/ phi = 0.9" = y1,
    "AR(1) w/ phi = 1" = y2,
    "AR(1) w/ phi = 1.01" = y3
  ) %>%
  tidyr::gather(model, y, -t) %>%
  ggplot(aes(x=t,y=y)) +
    geom_line() +
    facet_grid(model~., scale="free_y") 
```

## AR(1) - Negative $\phi$ 

```{r echo=FALSE, fig.height=5}
delta = 0.1
phi1 = 0.9
phi2 = 1.05

ar1 = data_frame(
  t = 1:200,
  y1 = 0,
  y2 = 0,
  y3 = 0
)

for(t in 2:nrow(ar1))
{
  ar1$y1[t] = delta - phi1 * ar1$y1[t-1] + rnorm(1)
  ar1$y2[t] = delta - ar1$y2[t-1] + rnorm(1)
  ar1$y3[t] = delta - phi2 * ar1$y3[t-1] + rnorm(1)
}

ar1 %>%
  rename(
    "AR(1) w/ phi = -0.9"  = y1,
    "AR(1) w/ phi = -1"    = y2,
    "AR(1) w/ phi = -1.01" = y3
  ) %>%
  tidyr::gather(model, y, -t) %>%
  ggplot(aes(x=t,y=y)) +
    geom_line() +
    facet_grid(forcats::as_factor(model)~., scale="free_y") 
```

## Stationarity of $AR(1)$ processes {.t}

Lets rewrite the AR(1) without any autoregressive terms


## Stationarity of $AR(1)$ processes {.t}

Under what conditions will an AR(1) process be stationary?



## Properties of a stationary $AR(1)$ processes {.t}



## Identifying AR(1) Processes

```{r echo=FALSE}
sims = data_frame(
  t = 1:100,
  "phi=-0.9" = arima.sim(n = 100, list(ar = -0.9)) %>% strip_attrs(),
  "phi=-0.5" = arima.sim(n = 100, list(ar = -0.5)) %>% strip_attrs(), 
  "phi= 0.5" = arima.sim(n = 100, list(ar =  0.5)) %>% strip_attrs(), 
  "phi= 0.9" = arima.sim(n = 100, list(ar =  0.9)) %>% strip_attrs()
)

sims %>%
  tidyr::gather(model, vals, -t) %>%
  ggplot(aes(x=t, y=vals)) +
    geom_line() +
    facet_wrap(~model)
```

## Identifying AR(1) Processes - ACFs

```{r echo=FALSE, fig.height=5}
with(
  sims,
  gridExtra::grid.arrange(
    forecast::ggAcf(`phi= 0.5`),
    forecast::ggAcf(`phi= 0.9`),
    forecast::ggAcf(`phi=-0.5`), 
    forecast::ggAcf(`phi=-0.9`)
  )
)
```

## Identifying AR(1) Processes - PACFs

```{r echo=FALSE, fig.height=5}
with(
  sims,
  gridExtra::grid.arrange(
    forecast::ggPacf(`phi= 0.5`),
    forecast::ggPacf(`phi= 0.9`),
    forecast::ggPacf(`phi=-0.5`), 
    forecast::ggPacf(`phi=-0.9`)
  )
)
```



