---
title: "Lecture 6" 
subtitle: "Discrete Time Series"
date: "9/21/2018"
fontsize: 11pt
output: 
  beamer_presentation:
    theme: metropolis
    highlight: pygments
    fig_caption: false
    latex_engine: xelatex
    includes:
      in_header: ../settings.tex
---
  
```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(forecast)
```

```{r config, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width=7,
  fig.height=4.5,
  out.width="\\textwidth",
  fig.align="center",
  echo=TRUE,
  warning=FALSE
)

ggplot2::theme_set(ggplot2::theme_bw())
```

# Discrete Time Series

## Stationary Processes {.t}

A stocastic process (i.e. a time series) is considered to be *strictly stationary* if the properties of the process are not changed by a shift in origin. 

. . .

In the time series context this means that the joint distribution of $\{y_{t_1}, \ldots, y_{t_n}\}$ must be identical to the distribution of $\{y_{t_1+k}, \ldots, y_{t_n+k}\}$ for any value of $n$ and $k$.



## Weak Stationary {.t}

Strict stationary is unnecessarily strong / restrictive for many applications, so instead we often opt for *weak stationary* which requires the following,

1. The process has finite variance 
$$E(y_t^2) < \infty \text{ for all $t$}$$

2. The mean of the process is constant 
$$E(y_t) = \mu \text{ for all $t$}$$

3. The second moment only depends on the lag 
$$Cov(y_t,y_s) = Cov(y_{t+k},y_{s+k}) \text{ for all $t,s,k$}$$

. . . 

When we say stationary in class we will almost always mean *weakly stationary*.


## Autocorrelation {.t}

For a stationary time series, where $E(y_t)=\mu$ and $\text{Var}(y_t)=\sigma^2$ for all $t$, we define the autocorrelation at lag $k$ as

$$
\begin{aligned}
\rho_k &= Cor(y_t, \, y_{t+k}) \\
       &= \frac{Cov(y_t, y_{t+k})}{\sqrt{Var(y_t)Var(y_{t+k})}} \\
       &= \frac{E\left( (y_t-\mu)(y_{t+k}-\mu) \right)}{\sigma^2}
\end{aligned}
$$

. . .

this is also sometimes written in terms of the autocovariance function ($\gamma_k$) as
$$
\begin{aligned}
\gamma_k &= \gamma(t,t+k) = Cov(y_t, y_{t+k}) \\
\rho_k &= \frac{\gamma(t,t+k)}{\sqrt{\gamma(t,t) \gamma(t+k,t+k)}} = \frac{\gamma(k)}{\gamma(0)}
\end{aligned}
$$

## Covariance Structure

Based on our definition of a (weakly) stationary process, it implies a covariance of the following structure,

\footnotesize

$$
\symbf{\Sigma} = \left(
\begin{matrix}
\gamma(0)   & \gamma(1)   & \gamma(2)   & \gamma(3)   & \cdots & \gamma(n-1) &\gamma(n)   \\
\gamma(1)   & \gamma(0)   & \gamma(1)   & \gamma(2)   & \cdots & \gamma(n-2) &\gamma(n-1) \\
\gamma(2)   & \gamma(1)   & \gamma(0)   & \gamma(1)   & \cdots & \gamma(n-3) &\gamma(n-2) \\
\gamma(3)   & \gamma(2)   & \gamma(1)   & \gamma(0)   & \cdots & \gamma(n-4) &\gamma(n-3) \\
\vdots      & \vdots      & \vdots      & \vdots      & \ddots & \vdots      & \vdots     \\
\gamma(n-1) & \gamma(n-2) & \gamma(n-3) & \gamma(n-4) & \cdots & \gamma(0)   & \gamma(1)  \\
\gamma(n)   & \gamma(n-1) & \gamma(n-2) & \gamma(n-3) & \cdots & \gamma(1)   & \gamma(0)  \\
\end{matrix}
\right)
$$



where $P_{t,k}(y)$ is the project of $y$ onto the space spanned by $y_{t+1},\ldots,y_{t+k-1}$.


## Example - Random walk {.t}

Let $y_t = y_{t-1} + w_t$ with $y_0=0$ and $w_t \sim \mathcal{N}(0,1)$.

```{r echo=FALSE}
rw = data_frame(
  t = 1:1000,
  y = cumsum(c(0, rnorm(999)))
)

ggplot(rw, aes(x=t, y=y)) + geom_line() + labs(title="Random walk")
```

## ACF + PACF

```{r echo=FALSE, fig.height=5}
forecast::ggtsdisplay(rw$y, lag.max = 50)
```

## Stationary? {.t}

Is $y_t$ stationary?





## Partial Autocorrelation - pACF

Given these type of patterns in the autocorrelation we often want to examine the relationship between $y_t$ and $y_{t+k}$ with the (linear) dependence of $y_t$ on $y_{t+1}$ through $y_{t+k-1}$ removed. 

This is done through the calculation of a partial autocorrelation ($\alpha(k)$), which is defined as follows:

$$
\begin{aligned}
\alpha(0) &= 1 \\
\alpha(1) &= \rho(1) = Cor(y_t,y_{t+1})\\
&~~\vdots \\
\alpha(k) &= Cor(y_t - P_{t,k}(y_t),~ y_{t+k} - P_{t,k}(y_{t+k}))
\end{aligned}
$$

## Example - Random walk with drift {.t}

Let $y_t = \delta + y_{t-1} + w_t$ with $y_0=0$ and $w_t \sim \mathcal{N}(0,1)$.

```{r echo=FALSE}
rwt = data_frame(
  t = 1:1000,
  y = cumsum(c(0, 0.1+rnorm(999)))
) 

ggplot(rwt, aes(x=t, y=y)) + geom_line() + labs(title="Random walk with trend")
```


## ACF + PACF

```{r echo=FALSE, fig.height=5}
forecast::ggtsdisplay(rwt$y, lag.max = 50)
```

## Stationary? {.t}

Is $y_t$ stationary?



## Example - Moving Average

Let $w_t \sim \mathcal{N}(0,1)$ and $y_t = w_{t-1}+w_t$.

```{r echo=FALSE, warning=FALSE}
ma = data_frame(
  t = 1:100,
  w = rnorm(100)
) %>%
  mutate(
    y = (c(NA,w[-100]) + w)
  )

ggplot(ma, aes(x=t, y=y)) + geom_line() + labs(title="Moving Average")
```

## ACF + PACF

```{r echo=FALSE, fig.height=4}
forecast::ggtsdisplay(ma$y, lag.max = 50, na.action = na.omit)
```

## Stationary? {.t}

Is $y_t$ stationary?


## Autoregressive

Let $w_t \sim \mathcal{N}(0,1)$ and $y_t = y_{t-1} - 0.9 y_{t-2} + w_t$ with $y_t = 0$ for $t < 1$.

```{r echo=FALSE, warning=FALSE}
ar = data_frame(
  t = 1:500,
  w = rnorm(500),
  y = NA
)
      
for(i in seq_along(ar$w))
{
  if (i == 1)
    ar$y[i] = ar$w[i]
  else if (i==2)
    ar$y[i] = ar$y[i-1] + ar$w[i]
  else
    ar$y[i] = ar$y[i-1] -0.9*ar$y[i-2] + ar$w[i]
}
  
ggplot(ar, aes(x=t, y=y)) + geom_line() + labs(title="Autoregressive")
```

## ACF + PACF

```{r echo=FALSE, fig.height=4}
forecast::ggtsdisplay(ar$y, lag.max = 50, na.action = na.omit)
```

## Example - Australian Wine Sales

Australian total wine sales by wine makers in bottles <= 1 litre. Jan 1980 – Aug 1994.

```{r}
aus_wine = readRDS("../data/aus_wine.rds")
aus_wine
```

## Time series

```{r echo=FALSE}
ggplot(aus_wine, aes(x=date, y=sales)) + 
  geom_line() + 
  geom_point()
```

## Basic Model Fit

```{r echo=FALSE}
l  = lm(sales ~ date, data=aus_wine)
l2 = lm(sales ~ date + I(date^2), data=aus_wine)

d = aus_wine %>%
  modelr::add_predictions(l, var="linear") %>%
  modelr::add_predictions(l2, var="quadratic")

d %>%
  select(-sales) %>%
  tidyr::gather(model, sales, -date) %>%
  ggplot(aes(x=date, y=sales, color=model)) + 
    geom_line(data=d, color="black") + 
    geom_point(data=d, color="black") +
    geom_line(size=1.5, alpha=0.75)
```

## Residuals

```{r echo=FALSE}
d = aus_wine %>%
  modelr::add_residuals(l, var="lin_resid") %>%
  modelr::add_residuals(l2, var="quad_resid")

d %>%
tidyr::gather(type, residual, -(date:sales)) %>%
ggplot(aes(x=date, y=residual, color=type)) + 
  geom_point() +
  geom_line() +
  facet_wrap(~type, nrow=2)
```




## Autocorrelation Plot

```{r echo=FALSE}
forecast::ggtsdisplay(d$quad_resid, lag.max = 36)
```

##

```{r echo=FALSE, message=FALSE, warning=FALSE}
d_lags = bind_cols(
  d,
  purrr::map_dfc(0:12, ~ list(lag = lag(d$quad_resid, n=.x))) %>% select(-lag)
) %>% 
  tidyr::gather(lag, lag_value, -(date:quad_resid)) %>%
  mutate(lag = forcats::as_factor(lag))

ggplot(d_lags, aes(x=lag_value, y=quad_resid)) +
  geom_point() +
  facet_wrap(~lag, ncol=4) +
  geom_smooth(method="lm", color='red', se = FALSE, alpha=0.1)
```




## Auto regressive errors

```{r echo=FALSE}
d_ar = mutate(d, lag_12 = lag(quad_resid, 12))
l_ar = lm(quad_resid ~ lag_12, data=d_ar)
summary(l_ar)
```

## Residual residuals

```{r echo=FALSE, warning=FALSE}
d_ar = d_ar %>%
  modelr::add_residuals(l_ar)

d_ar %>%
  ggplot(aes(x=date, y=resid)) +
  geom_point() + 
  geom_line()
```

## Residual residuals - acf

```{r echo=FALSE}
forecast::ggtsdisplay(l_ar$residuals, lag.max = 36)
```

##

```{r echo=FALSE, warning=FALSE, fig.height=5.5}
bind_cols(
  d_ar %>%
    modelr::add_residuals(l_ar) %>%
    select(-lag_12, -lin_resid, -quad_resid),
  purrr::map_dfc(0:12, ~ list(lag = lag(d_ar$resid, n=.x))) %>% select(-lag)
) %>%
  tidyr::gather(lag, lag_value, -(date:resid)) %>%
  mutate(lag = forcats::as_factor(lag)) %>%
  ggplot(aes(x=lag_value, y=resid)) +
    geom_point() +
    facet_wrap(~lag, ncol=4) +
    geom_smooth(method="lm", color='red', se = FALSE, alpha=0.1)
```

## Writing down the model? {.t}

So, is our EDA suggesting that we fit the following model?

$$ \text{sales}(t) = \beta_0 + \beta_1 \, t + \beta_2 \, t^2 + \beta_3 \, \text{sales}(t-12) + \epsilon_t $$

. . .

the model we actually fit is,

$$ \text{sales}(t) = \beta_0 + \beta_1 \, t + \beta_2 \, t^2 + w_t $$
where

$$ w_t = \delta \, w_{t-12} + \epsilon_t $$






